---
title: "Story 2: Do stricter gun laws reduce firearm gun deaths?"
author: "Naomi Buell"
format: pptx
editor: visual
---

```{r}
#| label: load packages
#| message: false
library(tidyverse)
library(janitor)
library(jsonlite)
library(RSocrata)
library(readxl)
library(maps)
library(RColorBrewer)
```

## Intro

The CDC publishes firearm mortality for each State per 100,000 persons <https://www.cdc.gov/nchs/pressroom/sosmap/firearm_mortality/firearm.htm>. Each State' firearm control laws can be categorized as very strict to very lax. The purpose of this Story is to answer the question, " Do stricter firearm control laws help reduce firearm mortality?"

For this assignment I:

-   Access the firearm mortality data from the CDC using an available API (<https://open.cdc.gov/apis.html>)
-   Create a 5 point Likert scale categorizing gun control laws from most lax to strictest and assign each state to the most appropriate Likert bin
-   Determine whether stricter gun control laws result in reduced gun violence deaths
-   Present my story using heat maps
-   Use an accessible color palette

## Get data through API

First, I get the firearm data, following instructions online: <https://dev.socrata.com/foundry/data.cdc.gov/489q-934x>.

```{r}
#| label: set up CDC API
#| error: true
#| message: false
#| warning: false
#| results: hide

# API credentials
base_url <- "https://data.cdc.gov/resource/489q-934x.json"
app_token <- "sx3mlnOPJgxWbgMEi9vT6hg9I"
spsmail <- "naomi.buell13@spsmail.cuny.edu"
password <- "Mi6P&ZPQt3eCRKQ3"

# API filters
cause_of_death <- "Firearm-related injury"
rate_type <-  "Age-adjusted"
time_period <- "12 months ending with quarter"

# Create URL, modeling this example: https://data.cdc.gov/resource/489q-934x.json?cause_of_death=All causes
composite_url <- paste0(base_url,
              "?cause_of_death=",
              cause_of_death,
              "&rate_type=",
              rate_type,
              "&time_period=",
              time_period)
```

Retrieve the data using the API and clean data frame. I elect to use quarterly age-adjusted death rates per 100,000 (calculated over the past 12 months) from firearm-related injuries.

```{r}
#| label: read firearms deaths and clean

df_cdc <- read.socrata(
  composite_url,
  app_token = app_token,
  email     = spsmail,
  password  = password
) |>
  select(-c(time_period, cause_of_death, rate_type, unit)) |>
  pivot_longer(
    starts_with("rate"),
    names_to = "state",
    values_to = "deaths",
    names_prefix = "rate_"
  ) |>
  clean_names() |>
  separate_wider_regex(year_and_quarter,
                       patterns = c(year = "[0-9]+", " Q", quarter = "[0-9]+")) |> 
  mutate(year = year |> as.numeric(),
         quarter = quarter |>  as.numeric(),
         deaths = deaths |> as.numeric(),
         state = str_replace_all(state, "_", " ")) |> 
  drop_na() |> 
  arrange(desc(year), desc(quarter), state)

head(df_cdc)
```

## Import and categorize gun control laws

I use data on the strictness of gun control laws from <https://giffords.org/lawcenter/resources/scorecard/>.

```{r}
#| label: import and clean gun control data
#| warning: false

gun_laws_path <- "https://raw.githubusercontent.com/naomibuell/DATA608/main/GiffordsScorecard.csv"
df_gun_laws <- read_csv(gun_laws_path) |> 
  clean_names() |> 
  mutate(state = str_to_lower(state)) |> 
  select(-starts_with("gun_death"))
  
head(df_gun_laws)
```

Then I merge the gun control data with the gun deaths data from the CDC.

```{r}
#| label: merge
#| warning: false

df_merged <- full_join(df_cdc, df_gun_laws)
```

Now that I have sourced all of my data, I create a 5 point Likert scale categorizing gun control laws from most lax to strictest and assign each state to the most appropriate Likert bin. The 5-point Likert scale will correspond with Giffords' grades of A, B, C, D, and F. I also do some final data type cleaning. I also add abbreviated state names.

```{r}
#| label: Likert

df_likert <- df_merged |>
  mutate(likert = substr(grade, 0, 1) |> # Adding likert scale
           factor(levels = c("A", "B", "C", "D", "F")),
         # Combining year and quarter as date) 
         date = as.Date(paste0(year, "-", (quarter - 1) * 3 + 1, "-01")),
         grade = as_factor(grade),
         state = as_factor(state),
         abbrev = state.abb[match(state, str_to_lower(state.name))] |> as_factor())

head(df_likert)
```

## Graph data

In this section, I produce heat maps and other graphs to check if stricter gun control laws result in reduced gun violence deaths, using accessible colors.

### Heat maps

Firearm mortality by state:

```{r}
#| label: Firearm Mortality by State

# Load map data
us_map <- map_data("state")

# Merge your data with the map data (make sure state names match and are lowercase)
df_map <- us_map |> 
  left_join(df_likert, by = c("region" = "state"))

# Plot heat map
ggplot(df_likert, aes(map_id = state, fill = deaths)) +
  geom_map(color = "darkgrey", map = us_map, linewidth = .3) +
  expand_limits(x = us_map$long, y = us_map$lat) +
  scale_fill_distiller(palette = "RdYlBu",
                       direction = -1) +  # Accessible palette
  theme_void() +  # Remove gridlines and background
  labs(fill = "Deaths per 100k")
```

Quintiles of deaths per 100k:

```{r}
#| label: plotting heatmap of deaths by quintile

df_quint <- df_likert |>
  drop_na(likert) |>
  filter(year == max(year), quarter == max(quarter)) |>
  mutate(
    quint = ntile(deaths, 5),
    # Create labels for each quintile with min/max deaths
    quint_label = factor(
      quint,
      levels = 1:5,
      labels = sapply(1:5, function(i) {
        min_max <- range(deaths[ntile(deaths, 5) == i])
        paste0(min_max[1], " - ", min_max[2], "")
      })
    )
  )

df_quint |> 
  ggplot(aes(map_id = state, fill = quint_label)) +
  geom_map(map = us_map, color = "gray", linewidth = .3) +
  expand_limits(x = us_map$long, y = us_map$lat) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1) +
  theme_void() +  # Removes gridlines and background
  labs(fill = "Deaths per 100k")

```

Gun regulation score by state:

```{r}
#| label: Gun regulation score by state
# Filter colorblind-friendly palettes
brewer.pal.info[brewer.pal.info$colorblind == TRUE, ]

# Plot the heat map using a discrete color scale
df_likert |> 
  drop_na(likert) |> 
  ggplot(aes(map_id = state, fill = likert)) +
  geom_map(map = us_map, color = "gray", linewidth = .3) +
  expand_limits(x = us_map$long, y = us_map$lat) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1) +
  theme_void() +  # Removes gridlines and background
  labs(fill = "Gun regulation grade")
```

### Bar chart with trendline

```{r}
#| label: scatter plot
#| warning: false

# Plot bars and trendline
df_likert |>
  # Choose only the latest quarter's data
  filter(year == max(year), quarter == max(quarter)) |>
  drop_na() |>
  ggplot(aes(y = deaths)) +
  # Bar chart
  geom_col(aes(x = fct_reorder(abbrev, gun_law_strength_ranked)), fill = brewer.pal(9, "RdYlBu")[3]) +
  # Trendline
  geom_smooth(
    aes(x = gun_law_strength_ranked),
    method = "lm",
    se = F,
    color = brewer.pal(9, "RdYlBu")[1]
  ) +
  # Theme adjustments
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(y = "Deaths per 100k", x = "States, ranked from most regulated to least regulated") +
  scale_x_discrete(breaks = function(x) x[seq(1, length(x), by = 49)])  # Show every 5th state label
```

```{r}
#| label: stats to call out

# Model number of deaths per 100k from state gun law ranking
lm_ranking <- lm(deaths ~ gun_law_strength_ranked, data = df_likert) 
lm_ranking |> summary()
# Model number of deaths per 100k from state grade
lm_likert <- lm(deaths ~ likert, data = df_likert) 
lm_likert |> summary()

df_likert |>
  # Choose only the latest quarter's data
  filter(year == max(year), quarter == max(quarter)) |>
  drop_na() |>
  arrange(desc(deaths)) |> 
  head(50)
```

For every rank you go down on the list of most regulated to least regulated states in terms of gun control laws, your state's death rate increases by .3 deaths per 100k per year. A perfect grade is associated with 9.6 gun deaths per 100k per year at baseline. A B adds an additional 3 deaths per 100k on top of that, a C would've added 7.6 deaths per 100k, a D would add 5.4 per 100k, and an F would create an additional 10.4 deaths per 100k. (All statistically significant based on linear regression). We can predict 43.99% of variance in gun deaths using gun law grading, and 47.24% of variance in gun deaths using gun law ranking.
